---
title: "Proteomics dataset EDA"
output: html_notebook
---

This .Rmd contains UbiLength EDA analysis with notes, supported by a guided workthrough (Arne Smits - Introduction to DEP).
UbiLength: an affinity enrichment dataset of ubiquitin interactors

Load UbiLength proteomics LC-MS dataset.

```{r}
library(DEP)
data("UbiLength")
#affinity enrichment of ubiquitin interactors
View(UbiLength)

library(tidyverse)
data <- UbiLength #dim = 3006,23

# remove "potential contaminants"
data <- filter(data, Reverse !="+", Potential.contaminant != "+") #dim = 2941,23

colnames(data)
# use LFQ.intensity columns for downstream analysis (label-free quantitation)
```


*Cleaning / normalising data*

```{r}
#duplicate genes?
#my ver:
u <- select(data, Gene.names) %>%
  unique()
dim(u)

#example ver:
data$Gene.names %>%
  duplicated() %>%
  any()
# i.e. of subgroup data$Gene.name, do duplicates exist? T/F

#exemplar table generated of unique Gene.names:
data %>% group_by(Gene.names) %>% # take subgroup
  summarise(frequency = n()) %>% #returns all samples(?)
  arrange(desc(frequency)) %>% filter(frequency > 1) #rearrange Gene.names in descending order of frequency (duplicates)

```

Identified duplicate genes are present in high frequencies (e.g. ATXN2 = 7x)
Rename each individual sample with "Gene.name;Protein.IDs":
```{r}
#exemplar:
data_unique <- make_unique(data, "Gene.names", "Protein.IDs", delim = ";") #make_unique from DEP package, call data > call column names within one protein group (row) > join character strings with delimiter ==> creates 2 new columns: names, ids

#confirm any duplicates left under new column 'names'?
data_unique$name %>% duplicated() %>% any()

```


Create summarised experiment (unique protein > LFQ intensity):
```{r}
LFQ_columns <- grep("LFQ.", colnames(data_unique)) #grab all column names that start with "LFQ."
head(LFQ_columns) #highlights relevant columns from column 9 onward
data_se_parsed <- make_se_parse(data_unique, LFQ_columns) #stitch unique proteins + data from relevant LFQ columns (9-14) into data.frame
#Assay data is automatically log2transformed. Listdata contains label, ID, replicates

```

Normalising dataset: some replicates in dataset are not quantified. Replicate = repeats; some return missing / NA / NaN values. Some only have data for one replicate (also extreme/unexpected) --> needs imputation (only if missing value are the minority, else risk false positive).
For proteins with excess missing values, do not attempt to impute. Remove/filter these proteins by setting a threshold:
```{r}
data_filt <- filter_missval(data_se_parsed, thr = 0) #take summarisedExperiment data.frame, filter proteins with 0 missing values (for at least 1 condition)
#each sample is replicated 3x (view data_uniqe; 3 LFQ intensity columns per condition)
#- above is strict (allows no missing values, filters proteins with all 3 replicates containing no missing data);

plot_numbers(data_filt) # barplot
#average no. of proteins per sample dropped from ~1900 to ~1500

plot_coverage(data_filt) #overlapping no. of proteins (?) across samples
```

For less strict filter: apply thr = 1 to allow max 1 missing value for any replicate
i.e. there are 3 replicates/samples for each condition. under thr = 1: if 2/3 replicates/samples have values in at least 1 condition, keep the row (even if there are 0/3 recorded values for another condition)



Normalise data via variance-stabilising transformation, to correct heavily skewed data. Variance independent of mean (vsn)
```{r}
data_norm <- normalize_vsn(data_filt)
#corrected unequal variance?:
plot_normalization(data_norm, data_filt)

```
Data distribution should be same/similar following normalisation.
Variance stabilising normalisation (VSN) stabilises/corrects unequal variance caused by low-abundance proteins with low variance and high-abundance proteins with high variance. (log fold change will only amplify the high-abundance proteins, hence not suitable)

Above data filtered under thr = 0: at least one condition must have 3/3 replicate/sample values.
Are there missing values for 1< condition for a particular protein? plot heatmap:
```{r}
plot_missval(data_filt)
```
Heat map: missing values = 0; not missing = 1
Based on the trend of missing/non-missing values, this could indicate if a protein was missing values at random (biological event) MAR, or errors from missing detection limit MNAR (not at random).



Choose the correct imputation method based on heatmap pattern:
1. There are missing values for several conditions: MAR?
2. Missing the detection limit threshold: MNAR?
3. .. therefore, are missing values biased to lower intensity proteins?:
```{r}
plot_detect(data_norm)
#density graph: a high proportion of TRUE missing values derive from low-intensity values ==> MNAR
#cmf graph: TRUE missing values at lower log fold intensities; FALSE missing value at higher intensities
```
High proportion of missing values at low intensities suggests below threshold limit ==> MNAR.
MNAR values imputed via left-censored imputation (impute feasible/estimated low values past threshold limit)

MNAR: quantile regression based left-censored function (QRILC), assess end tail of data (cut-off from limit threshold), apply linear regression with sampled data, fill end tail. In box plot, this would be reflected as end whisker (minimum value) uniform across all <LOD samples (bc used same sample to perform linear regression and predict missing <LOD values).
MAR: kNN (k-nearest neighbour) / maximum likelihood estimate (MLE) assessing overall trend to impute low intensity values
```{r}
#select imputation method:
#data_imp <- impute(data_norm, fun = "QRILC")
data_imp <- impute(data_norm, fun = "MinProb", q = 0.01) # use lowest 1% percentile of collected data for random sampling > impute values

plot_detect(data_imp)
plot_missval(data_imp)
#both return no missing values
```

```{r}
plot_imputation(data_norm, data_imp)
```




*Differential Expression Analysis*

```{r}
library(limma)
data_diff <- test_diff(data_imp, type = "control", control = "Ctrl_") # apply Bayes statistics to compare all samples against control (Ctrl_); i.e. is there a significant difference between control and each condition?

data_diff_all_contrasts <- test_diff(data_imp, type = "all") # compare and contrast all sample combinations (*spider leg friction coefficient table), i.e. is there a significant difference *between each condition* in protein expression?




```
Error: run test_diff() with a valid control.
Valid controls are: 'Ubi4_', 'Ubi6_', 'Ctrl_', 'Ubi1_'

To identify the correct name/character string, use df$condition.


(Each individual control replicate ("Ctrl__1", "Ctrl__2", "Ctrl__3") does not need to be specified. Use "Ctrl_" to group all 3 replicates for accurate df + sd + p-vals)

This leaves (beta) coefficients for each condition's replicates:
```{r}
#determine significant p-values (from control)
dep <- add_rejections(data_diff, alpha = 0.05, lfc = log2(1.5))

```


*Visualising the results*

PCA plot uses hierarchical clustering to group similar proteins together based on qualities, i.e. protein expression LFQ intensity

```{r}
plot_pca(dep)
#differences in x-axis is more significant than that of vertical y-axis difference
```
Ubi6 and Ubi4 are the most different in overall protein expression compared to Ctrl.
Ubi1 is similar in protein expression to Ctrl (PCA1)
#elliptic shape: wide length holds most variance (PCA1), short length has least variance (PCA2), therefore differences in PCA1 take precedence

This is further supported by a Pearson correlation heatmap:
```{r}
plot_cor(dep, lower = 0, pal = "RdPu")
#similar protein expression in Ubi6 and Ubi4 also reflected with a strong correlation in the heatmap
#weaker correlation in Ubi1 to Ctrl (strong PCA2, weak PCA1 difference)
```


Log fold intensity (log2 intensity) represents absolute abundance of a protein / expression; dataset is normalised (based on that protein's expression average z-score). Log2 intensity used so that proteins with exponentially higher expression do not occlude the least expressed proteins.

```{r}
#hierachical clustering heatmap (easier to view protein expression patterns across conditions/replicates)
plot_heatmap(dep, type = "centered", show_row_names = FALSE, col_limit= 4)

# use kmeans = TRUE to group protein clusters of rows (specified no. with k = x)
```
#kmeans = FALSE uses hierarchical clustering: bottom-up approach joining 2 most similar proteins, then cluster with next pair of similar proteins etc...


Log fold change lfc represents changes in protein expression from control (contrasts). Does not tell you by how much, only relative difference in expression.
```{r}
plot_heatmap(dep, type = "contrast", kmeans = TRUE, col_limit = 10, show_row_names = FALSE)
```
#kmeans = TRUE uses k-means clustering: define number of clusters with k = x, then centroids (each cluster's average) is chosen (distinct, least similar values to clearly define each cluster). Protein expression values most similar to a centroid will get regrouped. Each recruitment of a value recalculated the centroid average until all values have been assigned a cluster.


Volcano plots to visualise contrasts: changes in protein expression in one condition to another. x-axis = protein expression (contrast lfc), y-axis = p-value
```{r}
plot_volcano(dep, contrast = "Ubi6__vs_Ctrl_", label_size = 2)

#or select an individual protein:
plot_single(dep, proteins = "USP15", type = "centered")

#or visualise overall which proteins had significant protein expression + overlaps/similarities in proteins:
plot_cond(dep)
```

